Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Integrate all comments first part introduction

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Mu\\~{n}oz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. 
For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?]

Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. 
~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~ 
These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.
~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~ 
Multilevel models typically accommodate for this variability by including a separate group mean for each cluster. In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. [TD: emphasize when/why we need to account for clustering in the analysis of clustered data. Why is the presence of clustering relevant when considering multiple imputation of missing data? e.g. distinction between systematically and sporadically missing data. But also: mechanisms of missing data (e.g. MCAR, MAR, MNAR) may differ between clusters. But also: relation between observed data may differ between clusters? When/ why should we avoid using traditional imputation methods? e.g. congeniality issues.]

@@

ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black") +
  geom_text(aes(label = text), color = "black") +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(breaks = 1:6, labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])), name = NULL, position = "top") +
  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
  theme_minimal() +
  theme(panel.grid = element_blank())

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

In many contemporary data analysis efforts, some form of hierarchical or clustered data structures are recorded. In the simplest case, such a structure entails the nesting of units within clusters (e.g., students within school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., data from diverse settings and populations within large databases). The clustered structure of multilevel data should be taken into account when developing analysis models: 1) for the simple reason that groups of observations share some common variance, and 2) because ignoring multilevel structures can be harmful to the statistical inferences and introduce bias in estimators [@hox17]. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. Table \ref{tab:def} provides an overview of some key concepts in multilevel modeling.

<!-- Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?] -->

<!-- Ignoring the clustered structure of such multilevel data can be harmful to the statistical inferences and introduce bias in estimators [@hox17].  -->
<!-- ~~Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations--and vice versa for each of the units that make up the cluster.~~  -->
<!-- These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance. -->
<!-- ~~The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.~~  -->
<!-- Multilevel models typically accommodate for variability by including a separate group mean for each cluster. In addition to random intercepts, multilevel models can also include random effects and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21].  -->

<!-- There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. [TD: emphasize when/why we need to account for clustering in the analysis of clustered data. Why is the presence of clustering relevant when considering multiple imputation of missing data? e.g. distinction between systematically and sporadically missing data. But also: mechanisms of missing data (e.g. MCAR, MAR, MNAR) may differ between clusters. But also: relation between observed data may differ between clusters? When/ why should we avoid using traditional imputation methods? e.g. congeniality issues.] -->

\begin{table}[tb]
\caption{Concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black") +
  geom_text(aes(label = text), color = "black") +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(breaks = 1:6, labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])), name = NULL, position = "top") +
  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Integrate all comments first part introduction

\author{
Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde\{n\}oz
Avila\\University Medical Center Utrecht \AND Valentijn de
Jong\\University Medical Center Utrecht \And Gerko Vink\\Utrecht
University \AND Thomas Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde\{n\}oz
Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

\sout{In many contemporary data analysis efforts, some form of
hierarchical or clustered structure is recorded.} Contemporary
decision-making is increasingly often data-driven and sometimes even
entirely dictated by data. Although data relevant for decision makers
were initially collected using small and well-designed studies, there
has been a growing need to address more complex questions at a wider
scale. This not only requires to collect larger amounts of data, but
also from more diverse settings and populations. For example, {[}TD:
introduce IPDMA here; eg. HIV studies Johanna is referring to{]}.
Another example is the use of large databases with information from
thousands or even millions of individuals from multiple cities, regions
or even countries. Finally, a third situation arises when data are
collected from multicenter studies {[}TD: maybe refer to the
schools/classes example here{]}. For example, students may be clustered
in classes in psychometrics research, or patients may be clustered in
studies in individual patient data meta-analyses in medical research. A
common characteristic in aforementioned examples is the presence of
clustering. {[}TD: briefly explain what is clustering?{]}

Ignoring the clustered structure of such multilevel data can be harmful
to the statistical inferences and introduce bias in estimators
\citep{hox17}. \sout{Imagine a case where cross-level interactions
between unit-level variables and cluster-level variables are present.
The cluster to which a unit belongs may then influence the unit-level
observations--and vice versa for each of the units that make up the
cluster.} These relations can and should be taken into account when
developing analysis models for multilevel data for the simple reason
that groups of observations share some common variance. \sout{The
variability due to clustering is often measured by means of the
intraclass coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure.} Multilevel models typically accommodate for this variability
by including a separate group mean for each cluster. In addition to
random intercepts, multilevel models can also include random effects and
heterogeneous residual error variances across clusters \citep[see
e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are many names
for models that take clustering into account. Some popular examples are
`multilevel models', `hierarchical models', `mixed effect models' and
`random effect models'. {[}TD: emphasize when/why we need to account for
clustering in the analysis of clustered data. Why is the presence of
clustering relevant when considering multiple imputation of missing
data? e.g.~distinction between systematically and sporadically missing
data. But also: mechanisms of missing data (e.g.~MCAR, MAR, MNAR) may
differ between clusters. But also: relation between observed data may
differ between clusters? When/ why should we avoid using traditional
imputation methods? e.g.~congeniality issues.{]}

>>>>

\author{
Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
Medical Center Utrecht \AND Valentijn de Jong\\University Medical
Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

In many contemporary data analysis efforts, some form of hierarchical or
clustered data structures are recorded. In the simplest case, such a
structure entails the nesting of units within clusters (e.g., students
within school classes). More complex clustered structures may occur when
there are multiple hierarchical levels (e.g., patients within hospitals
within regions or countries), or when the clustering is non-nested
(e.g., data from diverse settings and populations within large
databases). The clustered structure of multilevel data should be taken
into account when developing analysis models: 1) for the simple reason
that groups of observations share some common variance, and 2) because
ignoring multilevel structures can be harmful to the statistical
inferences and introduce bias in estimators \citep{hox17}. There are
many names for models that take clustering into account. Some popular
examples are `multilevel models', `hierarchical models', `mixed effect
models' and `random effect models'. Table \ref{tab:def} provides an
overview of some key concepts in multilevel modeling.

\begin{table}[tb]
\caption{Key concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}
